Software Architectures to Integrate Workflow Engines in Science Gateways

### 2016-04-27 paper.tex
Bold other names
For sub-tasking, Global Complexity and other small changes.

See Figure~\subref*{archi:service}. The workflow engine is available
externally to the science gateway, in a service. The science gateway
controls the service through a specific interaction (\texttt{d}) that
might be implemented as a web-service call (e.g., RESTful or SOAP), as
a command-line or as any other method that offers a well-defined
interface to the workflow engine. The workflow engine might be invoked
either as a black box that completely masks the infrastructure and
tasks, or as a white box that allows for some interaction with
them. The workflow engine is responsible for controlling the tasks on
the infrastructure (\texttt{b}) and for performing the required data
transfers to execute them (\texttt{c$_2$}). User data is usually
managed through the science gateway (\texttt{c$_1$}), although it
might as well be delivered by the workflow engine directly to the
user.

@@

\subsection{Sub-tasking}

@@

The CBRAIN-PSOM integration~\cite{GLAT-16} is shown on
Figure~\ref{fig:cbrain-psom-architecture}. The PSOM workflow engine
adopts a pilot-job architecture~\cite{turilli2015comprehensive} where
a master coordinates workflow execution by submitting workers and
establishing direct communication channels with them. Note how this
peculiar execution model is totally supported by the sub-tasking
architecture.

@@

The pool model was implemented in the SHIWA pool~\cite{ROGE-13}
diagrammed in Figure~\ref{fig:shiwa-pool-architecture}. In this
Figure, interaction \texttt{d} of Figure~\subref*{archi:agent} is
implemented in 3 different types of calls to the pool: workflow
submission (\texttt{1} \& \texttt{2}), workflow status retrieval
(\texttt{0} \& \texttt{3})\closednote{sorina}{I suppose letters come here
from the initial pool-model figure; it's a bit puzzling when mixed with
letters from Fig 2}{Tristan}{Fixed.}, and workflow retrieval (\texttt{13} \&
\texttt{14}). In Figure~\ref{fig:shiwa-pool-architecture},
interaction \texttt{f} of Figure~\subref*{archi:agent} is implemented
through 2 types of calls: workflow instance retrieval (\texttt{4},
\texttt{5} and \texttt{6}), and workflow instance update (\texttt{11}
and \texttt{12}). Workflow instance retrieval is used by the agents to
fetch work from the pool. Workflow instance update is used by the
agents to update workflow statuses. In the SHIWA pool, agents can wrap
different types of workflow engines to execute workflows expressed
with different languages. Calls \texttt{0}, \texttt{7}, \texttt{8} and
\texttt{9} on Figure~\ref{fig:shiwa-pool-architecture} are used by
workflow engine plugins to declare their supported language and launch
engines, and by workflow engines to report their status to the
agent. These calls are specific to the SHIWA Pool implementation of
the \texttt{agent} component and therefore have no corresponding
representation in Figure~\subref*{archi:agent}.

@@

\emph{Global complexity} is obtained by counting the total number of
interactions and components on the architecture diagram. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}), and
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$} since it is usually not developed by the groups who
integrate workflow engines in science gateway. However, integrating an
infrastructure in the architecture usually requires some technical
effort (e.g., account creation, software installation, etc.), which is
why we keep it in the metric. The two metrics are summed to obtain the
total score for global complexity. Global complexity measures the
total number of software pieces to develop, including interactions and
components. It may be used to measure the development effort required
to build the architecture.

@@

\paragraph{Extensibility} Integrating a new type of workflow engine
requires to develop interaction \texttt{e} and to install the engine
on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
requires updates only on the infrastructure (\texttt{E$_2$}=1).  New
workflows are integrated by creating a new task in the science gateway
through interactions \texttt{a$_1$} and \texttt{a$_2$}  (\texttt{E$_3$}=2). Adding a new
infrastructure requires to update interactions \texttt{a$_1$}, \texttt{b} and
\texttt{c} in the science gateway (\texttt{E$_4$}=3).

@@

\paragraph{Global complexity} Setting up a nested workflow
architecture with service invocation requires a science gateway, 2
workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
architecture involves 11 interactions (\texttt{I$_2$=11}):
\texttt{a$_1$} (appears twice: once for each infrastructure),
\texttt{a$_2$}, \texttt{b} (twice: once for each infrastructure),
\texttt{c$_1$} (twice: once for each infrastructure), \texttt{c$_2$}
(twice: once for each workflow engine and infrastructure) and
\texttt{d} (twice: once for each workflow engine).
\closedanswerednote{Rafael}{I like the idea of what has been discussed in
  the `integration effort', however I think the name may lead to
  misinterpretations. For example, according to Table 2, it requires
  more effort to integrate in a Tight architecture than in a
  service-based architecture. Considering the principles of software
  engineer, it is simpler to integrate/maintain a system using
  decoupled services. Maybe it should be renamed to something related
  to 'dependencies'.}{Tristan}{In our scope I'm not sure that it's
  simpler to integrate a service architecture than a tightly
  integrated one. But I take your point. 'Dependencies' is not very
  suitable either. What about 'development effort'?}{Tristan}{Replaced
  'Integration effort' with 'global complexity', as discussed in
  github issue \#8.}

>>>>

See Figure~\subref*{archi:service}. The workflow engine is available
externally to the science gateway, as a service. The science gateway
controls the service through a specific interaction (\texttt{d}) that
might be implemented as a web-service call (e.g., RESTful or SOAP), as
a command-line or as any other method that offers a well-defined
interface to the workflow engine. The workflow engine might be invoked
either as a black box that completely masks the infrastructure and
tasks, or as a white box that allows for some interaction with
them. The workflow engine is responsible for controlling the tasks on
the infrastructure (\texttt{b}) and for performing the required data
transfers to execute them (\texttt{c$_2$}). User data is usually
managed through the science gateway (\texttt{c$_1$}), although it
might as well be delivered by the workflow engine directly to the
user.

@@

\subsection{Sub-tasking} \note{Marc-e}{May I propose another name here? Either
"Task-centric" or "Task Oriented". My reasoning is that the WE is just another
task, as well as the sub-tasks themselves.}

@@

The CBRAIN-PSOM integration~\cite{GLAT-16} is shown on
Figure~\ref{fig:cbrain-psom-architecture}. The PSOM workflow engine
adopts a pilot-job architecture~\cite{turilli2015comprehensive} where
a master coordinates workflow execution by submitting workers and
establishing direct communication channels with them. Note how this
peculiar execution model is well supported by the sub-tasking
architecture.

@@

The pool model was implemented in the SHIWA pool~\cite{ROGE-13}
diagrammed in Figure~\ref{fig:shiwa-pool-architecture}. In this
Figure, interaction \texttt{d} of Figure~\subref*{archi:agent} is
implemented in 3 different types of calls to the pool: workflow
submission (\texttt{1} \& \texttt{2}), workflow status retrieval
(\texttt{0} \& \texttt{3})\closednote{sorina}{I suppose letters come here
from the initial pool-model figure; it's a bit puzzling when mixed with
letters from Fig 2}{Tristan}{Fixed.}, and workflow retrieval (\texttt{13} \&
\texttt{14}). \note{Marc-e}{Maybe easier to follow if the following would be in fig
 legend instead of here, as for 2 previous figures} In Figure~\ref{fig:shiwa-pool-architecture},
interaction \texttt{f} of Figure~\subref*{archi:agent} is implemented
through 2 types of calls: workflow instance retrieval (\texttt{4},
\texttt{5} and \texttt{6}), and workflow instance update (\texttt{11}
and \texttt{12}). Workflow instance retrieval is used by the agents to
fetch work from the pool. Workflow instance update is used by the
agents to update workflow statuses. In the SHIWA pool, agents can wrap
different types of workflow engines to execute workflows expressed
with different languages. Calls \texttt{0}, \texttt{7}, \texttt{8} and
\texttt{9} on Figure~\ref{fig:shiwa-pool-architecture} are used by
workflow engine plugins to declare their supported language and launch
engines, and by workflow engines to report their status to the
agent. These calls are specific to the SHIWA Pool implementation of
the \texttt{agent} component and therefore have no corresponding
representation in Figure~\subref*{archi:agent}.

@@

\emph{Global complexity} is obtained by counting the total number of
interactions and components on the architecture diagram. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}), and
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$} since it is usually not developed by the groups who
integrate workflow engines in science gateway. However, integrating an
infrastructure in the architecture usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), which is
why we keep it in the metric. The two metrics are summed to obtain the
total score for global complexity. Global complexity measures the
total number of software pieces to develop, including interactions and
components. It may be used to measure the development effort required
to build the architecture.

@@

\paragraph{Extensibility} Integrating a new type of workflow engine
requires to develop interaction \texttt{e} and to install the engine
on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
in this architechture shares the same mechanism as version updates of existing
tasks or tools, which requires an update on the infrastructure (\texttt{E$_2$}=1).  New
workflows are integrated by creating a new task in the science gateway
through interactions \texttt{a$_1$} and \texttt{a$_2$}  (\texttt{E$_3$}=2). Adding a new
infrastructure requires to update interactions \texttt{a$_1$}, \texttt{b} and
\texttt{c} in the science gateway (\texttt{E$_4$}=3).

@@

\paragraph{Global complexity} Setting up a nested workflow
architecture with service invocation requires a science gateway, 2
workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
architecture involves 11 interactions (\texttt{I$_2$=11}):
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c$_1$} (twice: once for each infrastructure),
\texttt{c$_2$} (twice: once for each workflow engine and infrastructure) and
\texttt{d} (twice: once for each workflow engine).
\closedanswerednote{Rafael}{I like the idea of what has been discussed in
  the `integration effort', however I think the name may lead to
  misinterpretations. For example, according to Table 2, it requires
  more effort to integrate in a Tight architecture than in a
  service-based architecture. Considering the principles of software
  engineer, it is simpler to integrate/maintain a system using
  decoupled services. Maybe it should be renamed to something related
  to 'dependencies'.}{Tristan}{In our scope I'm not sure that it's
  simpler to integrate a service architecture than a tightly
  integrated one. But I take your point. 'Dependencies' is not very
  suitable either. What about 'development effort'?}{Tristan}{Replaced
  'Integration effort' with 'global complexity', as discussed in
  github issue \#8.} \note{Marc-e}{I understand both points. However, complexity is
  addressed in Robustness as well... I liked the idea of development effort. It's a tough call.}
